\vspace{-1.5em} \bf A general and efficient representation of ancestral recombination graphs

### 2024-03-20 paper.tex
Edits to Section 3 (eARGs)

\section{Event ARGs}
\label{sec-eARG}
% Trying to be totally, boringly, abundantly, blatantly clear here what we're
% talking about.
In this section we define the classical view of an ARG data structure,
and illustrate its limitations. We are
interested in the details of how ARGs are described mathematically,
and as a consequence, how they are represented in a practical sense
as the output of inference programs.
Where details of an ARG data structure (the encoding) are
% Three papers 10 years apart will do here, right?
provided~\citep[e.g.][]{
wiuf1999recombination,gusfield2014recombinatorics,hayman2023recoverability}
they follow the approach described by Griffiths and colleagues
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions), and a large number of
% Note the "an" here to avoid cases like ARGweaver
ARG inference methods use it as an output format
\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
heine2018bridging,ignatieva2021kwarg}.
% Griffiths never mentions "sampling events", so let's not get
% into that. It's not an important detail, and we don't want to be
% overly precise about something that people haven't actually written
% down.
In this Griffiths encoding we have two types of internal node in the graph,
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints, and these breakpoints
are used to construct the local trees.
This is done by tracing pastwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}.
Because it is focused on recording events and their properties,
we will refer to this Griffiths encoding as the ``event ARG'' or eARG
encoding. Fig.~\ref{fig-event-arg} shows an example of a classical
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
Assigning a breakpoint to a recombination node is
not sufficient to uniquely define the local trees, and either
some additional ordering rules~\citep[e.g.][]{griffiths1996ancestral} or
explicit information~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}
is required to distinguish the left and right parents.
We assume in Fig.~\ref{fig-event-arg} that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

@@

% eARGs are representationally for limited interchange
While the Griffiths approach of annotating recombination nodes with a
breakpoint in an eARG is a concise and elegant way of describing realisations
of the coalescent, it is inherently limited when
implemented literally. The eARG encoding explicitly models only
two different types of event and thus anything that is not a single crossover
recombination or common ancestor event, must be incorporated
either in a roundabout way using these
events, or by adding new types of event to the encoding. For example, gene
conversion could be accommodated either by stipulating a third type of event
(annotated by two breakpoints and corresponding traversal conventions for
recovering the local trees) or by two recombination nodes joined by a
zero-length edge. From the perspective of practical interchange of data between
inference methods and downstream applications, both
workarounds are problematic, and the gARG encoding described in the previous
section offers a much simpler solution.

Aside from these obvious practical challenges arising from a
literal implementation of the Griffiths approach, there is a deeper
issue with the implicit strategy of basing an ARG data structure on
recording events and their properties (e.g.\ the crossover breakpoint
for a recombination event). The fundamental problem
is that this approach assumes all events are \emph{knowable}, and does not
provide any obvious mechanism for either aggregating multiple events
or expressing uncertainty about them. While this is not a
problem when describing the results of simulations (where all details
are perfectly known), it is a major issue when we wish to
formally describe the output of inference methods, particularly
as datasets approach the population scale.
As discussed in the introduction,
the precise details of all events in these vast ARGs can never be
known, and a data structure that \emph{enforces} complete
precision is therefore an impediment to progress.

There is also a certain clarity gained by explicitly modelling nodes
in the inheritance graph as genomes.
Outside of the context of a
% The point being, events are perfectly well defined in the models
% but not in reality
mathematical model, an ``event'' is a slippery concept.
For example, \emph{which} genome along a chain of cell divisions should be
regarded as the one where an event occurred,
or whether multiple coalescences
within a single individual should be regarded as one or multiple events are
debatable points (Appendix~\ref{sec-cell-lineages-and-args}).
% If we want to have an ARG software ecosystem then such fireside
% discussions won't help.
From the perspective of a concrete data structure,
ideally forming the basis of an ecosystem of interoperable
inference and analysis methods, such debates
are unproductive.

>>>>

\section{Event ARGs}
\label{sec-eARG}
In this section we define a common (and classical) view of an ARG data structure,
described explicitly in several publications~\citep[e.g.][]{
wiuf1999recombination,gusfield2014recombinatorics,hayman2023recoverability}
(but see~\citet{parida2011minimal} and \citet{zhang2023biobank}
for notable exceptions), and used as an output format by multiple ARG inferences tools
\citep[e.g.][]{song2004minimum,song2005efficient,rasmussen2014genome,
heine2018bridging,ignatieva2021kwarg}.
% Griffiths never mentions "sampling events", so let's not get
% into that. It's not an important detail, and we don't want to be
% overly precise about something that people haven't actually written
% down.
In this encoding we have two types of internal node in the graph,
representing the common ancestor and recombination events
in the history of a sample.
At common ancestor nodes, the inbound lineages merge into a
single ancestral lineage with one parent, and at recombination
nodes a single lineage is split into two independent
ancestral lineages. Recombination nodes are annotated with
the corresponding crossover breakpoints, and these breakpoints
are used to construct the local trees.
This is done by tracing pastwards through the graph from the samples,
making decisions about which outbound edge to follow through
recombination nodes based on the breakpoint
position~\citep{griffiths1996ancestral}.
Because it is focused on recording events and their properties,
we will refer to this encoding as the ``event ARG'' or eARG
encoding. Fig.~\ref{fig-event-arg} shows an example of an
eARG with three sample genomes (\noderef{a}, \noderef{b}, and \noderef{c}),
three common ancestor events (\noderef{e}, \noderef{f}, and \noderef{g})
and a single recombination event (node \noderef{d}) with a breakpoint
at position $x$.
Assigning a breakpoint to a recombination node is
not sufficient to uniquely define the local trees, and either
some additional ordering rules~\citep[e.g.][]{griffiths1996ancestral} or
explicit information~\citep[e.g.][]{gusfield2014recombinatorics,ignatieva2021kwarg}
is required to distinguish the left and right parents.
We assume in Fig.~\ref{fig-event-arg} that \noderef{d} inherits genetic material to the
left of $x$ from \noderef{e} and to the right of $x$ from \noderef{f}.

@@

% eARGs are representationally for limited interchange
While this approach of annotating recombination nodes with a
breakpoint in an eARG is a concise and elegant way of describing realisations
of the coalescent, it is inherently limited when
implemented literally. The eARG encoding explicitly models only
two different types of event and thus anything that is not a single crossover
recombination or common ancestor event, must be incorporated
either in a roundabout way using these
events, or by adding new types of event to the encoding. For example, gene
conversion could be accommodated either by stipulating a third type of event
(annotated by two breakpoints and corresponding traversal conventions for
recovering the local trees) or by two recombination nodes joined by a
zero-length edge. The gARG encoding described in the previous
section offers a simpler and more direct solution.

Aside from these practical challenges, there is also a deeper
issue with the implicit strategy of basing an ARG data structure on
recording events and their properties (e.g.\ the crossover breakpoint
for a recombination event). This approach assumes all events are 
\emph{knowable}, and does not
provide an obvious mechanism for either aggregating multiple events
or expressing uncertainty about them. While this is not a
problem when describing the results of simulations (where all details
are perfectly known), it is an issue when we wish to
formally describe the output of inference methods, particularly
as datasets approach the population scale.
As discussed in the introduction,
the precise details of all events in these vast ARGs can never be
known, so it is disadvantageous to use a data structure that enforces complete
precision.

There is also a certain clarity gained by explicitly modelling nodes
in the inheritance graph as genomes.
Outside of the context of a
% The point being, events are perfectly well defined in the models
% but not in reality
mathematical model, an ``event'' is a slippery concept.
For example, \emph{which} genome along a chain of cell divisions should be
regarded as the one where an event occurred,
or whether multiple coalescences
within a single individual should be regarded as one or multiple events are
debatable points (Appendix~\ref{sec-cell-lineages-and-args}).
% If we want to have an ARG software ecosystem then such fireside
% discussions won't help.
The gARG format avoids the need for such considerations,
so is more naturally compatible with forming an ecosystem of interoperable
inference and analysis methods.
